'''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''
Project_Synopsis
Data is analysed for storm and severe weather effects
on society. Two folds objective: first, recognize the
eventsdetrimental to public health and second, 
recognizethose which have most severe economic
consequences. The data is taken from NOAA Strom
database (1950-2011)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
```
# Libraries are imported as
```{r import_libs, echo=TRUE}
library(dplyr)
plot <- function(names, columns, totals, colors, main){
  colnames(totals) <- names
  
  par(las=2,mar=c(6,4,1,1))
  barplot(totals, col=colors,main=main,cex.names  = 0.8,cex.axis = 0.8)
  legend("topright", columns,fill=colors,bty = "n")
}
```

# Processing data as
```{r read_data, cache=TRUE}
StormData <- read.csv("data/StormData.csv.bz2")
colnames(StormData)
```

# Scrubbing the labels
```{r CleanEventType}
event_types <- as.data.frame(table(StormData$EVTYPE))
event_types <- event_types[order(event_types$Var1), ]
```
# Those records that possesses multiple events need to be properly counted and categorized
StormData$EVTYPE <- as.character(StormData$EVTYPE)
StormData$EVTYPE[grepl("/|&|and", StormData$EVTYPE,ignore.case = TRUE)] <- "Multiple Event"
StormData$EVTYPE[grepl("volc", StormData$EVTYPE,ignore.case = TRUE)] <- "Volcano"
StormData$EVTYPE[grepl("wind|wnd", StormData$EVTYPE,ignore.case = TRUE)] <- "WIND"
StormData$EVTYPE[grepl("funnel|tornado", StormData$EVTYPE,ignore.case = TRUE)] <- "Tornado"
StormData$EVTYPE[grepl("glaze", StormData$EVTYPE,ignore.case = TRUE)] <- "Glaze"
StormData$EVTYPE[grepl("hail", StormData$EVTYPE,ignore.case = TRUE)] <- "Hail"
StormData$EVTYPE[grepl("dust", StormData$EVTYPE,ignore.case = TRUE)]  <- "DUST"
StormData$EVTYPE[grepl("flood", StormData$EVTYPE,ignore.case = TRUE)] <- "FLOOD"
StormData$EVTYPE[grepl("ic(e|y)", StormData$EVTYPE,ignore.case = TRUE)] <- "Ice"
StormData$EVTYPE[grepl("fire|smoke", StormData$EVTYPE,ignore.case = TRUE)] <- "FIRE"
StormData$EVTYPE[grepl("thunder", StormData$EVTYPE,ignore.case = TRUE)] <- "Thunder Storm"
StormData$EVTYPE[grepl("slide|eros", StormData$EVTYPE,ignore.case = TRUE)] <- "Erosion"
StormData$EVTYPE[grepl("rain", StormData$EVTYPE,ignore.case = TRUE)] <- "Rain"
StormData$EVTYPE[grepl("freez|cold|snow|chill|winter", StormData$EVTYPE,ignore.case = TRUE)] <- "Cold Weather"
StormData$EVTYPE[grepl("TROPICAL.STORM", StormData$EVTYPE,ignore.case = TRUE)] <- "TROPICAL STORM"
StormData$EVTYPE[grepl("heat", StormData$EVTYPE,ignore.case = TRUE)] <- "Heat"
StormData$EVTYPE[grepl("(hurri|opal)", StormData$EVTYPE,ignore.case = TRUE)] <- "Hurricane"
```
# Classifying data to those which are relevant to the question
```{r seperate data}
health <- StormData[,(c(8,23:24))]
property<-StormData[,c(8,25:28)]
```
# Empty fields are replaced with magnitude zero
```{r Load Property Data}
table(property$PROPDMGEXP)
table(property$CROPDMGEXP)
property$PROPDMGEXP<-factor(property$PROPDMGEXP,levels=c("H","K","M","B","h","m","O"))
property$PROPDMGEXP[is.na(property$PROPDMGEXP)] <- "O"
property$CROPDMGEXP<-factor(property$CROPDMGEXP,levels=c("K","M","B","k","m","O"))
property$CROPDMGEXP[is.na(property$CROPDMGEXP)] <- "O"
```

# Compensating damage with the appropriate multiplier
 o(one) = 1
 h(undred)=100
 k(thousand)=1000
 m(million)=1000000
 b(billion)=1000000000

```{r Adjusting Damage Totals}
property$PROPDMGEXP <- as.character(property$PROPDMGEXP)
property$CROPDMGEXP <- as.character(property$CROPDMGEXP)
property$PROPDMGMLT <- 0
property$CROPDMGMLT <- 0
property$PROPDMGMLT[grepl("h", property$PROPDMGEXP,ignore.case = TRUE)]<-100
property$PROPDMGMLT[grepl("k", property$PROPDMGEXP,ignore.case = TRUE)]<-1000
property$PROPDMGMLT[grepl("m", property$PROPDMGEXP,ignore.case = TRUE)]<-1000000
property$PROPDMGMLT[grepl("b", property$PROPDMGEXP,ignore.case = TRUE)]<-1000000000
property$PROPDMGMLT[grepl("o", property$PROPDMGEXP,ignore.case = TRUE)]<-1
property$CROPDMGMLT[grepl("k", property$CROPDMGEXP,ignore.case = TRUE)]<-1000
property$CROPDMGMLT[grepl("m", property$CROPDMGEXP,ignore.case = TRUE)]<-1000000
property$CROPDMGMLT[grepl("b", property$CROPDMGEXP,ignore.case = TRUE)]<-1000000000
property$CROPDMGMLT[grepl("o", property$CROPDMGEXP,ignore.case = TRUE)]<-1
property$PROPDMG <- property$PROPDMG * property$PROPDMGMLT
property$CROPDMG <- property$CROPDMG * property$CROPDMGMLT
property$total <- property$PROPDMG + property$CROPDMG
```
'''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''RESULTS''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''''
# 1. Health Totals
```{r Population Health Total}
health.totals <- aggregate(cbind(FATALITIES,INJURIES) ~ EVTYPE, data = health, sum, na.rm=TRUE)
health.totals$TOTAL <- health.totals$FATALITIES + health.totals$INJURIES
health.totals <- health.totals[order(-health.totals$TOTAL), ]
health.totals <- health.totals[1:25,]
plot(health.totals$EVTYPE,
     as.matrix(t(health.totals[,c(-1,-4)])),
     colors = c("dark blue","red"),
     columns = c("Fatalities","Injuries"),
     main = "Most Harmful Event Type in the United States")
```
# 2. Population Health Assessment
Tornado is the largest contributor to the severity in the public health.
This however is not conclusive and require further investigation.
# 3. Economic Impact 
```{r Economic Total}
economic.total <- aggregate(cbind(PROPDMG,CROPDMG, total) ~ EVTYPE, data = property, sum, na.rm=TRUE)
economic.crop <- economic.total[order(-economic.total$CROPDMG), ]
economic.crop <- economic.crop[1:25,]
economic.prop <- economic.total[order(-economic.total$PROPDMG), ]
economic.prop <- economic.prop[1:25,]
plot(economic.prop$EVTYPE,
     as.matrix(t(economic.prop[,c(-1,-3,-4)])),
     colors = c("dark blue","red"),
     columns = c("Property Damage"),
     main = "Economic Impact of Weather on Propery Damage")
plot(economic.crop$EVTYPE,
     as.matrix(t(economic.crop[,c(-1,-2,-4)])),
     colors = c("dark blue","red"),
     columns = c("Crop Damage"),
     main = "Economic Impact of Weather on Crop Damage")
```     
# 4. Economic Impact Assessment
Drought has the largest effect on the crop whereas the flood contributes 
most severe impact on the public health.
